---
title: "The simdata dataset"
author: 
   - Matthias Kormaksson, Kostas Sechidis
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simdata-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ref.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The knockofftools package comes with a simulated data set called `simdata`. This short vignette demonstrates how the data set was generated.

```{r setup, message=FALSE}
library(knockofftools)
```

## Data generation with `generate_simdata` {#simdata}

The function `generate_simdata` was used to generate the data set `simdata` that comes with the R-package. 
```{r}
generate_simdata <- function() {

  RNGkind("L'Ecuyer-CMRG")
  set.seed(56969)

  N <- 2000
  p <- 30
  p_b = 10
  p_nn <- 10

  # Generate a 2000 x 30 Gaussian data.frame under equi-correlation(rho=0.5) structure,
  # with 10 of the columns dichotomized
  X <- generate_X(n=N, p=p, p_b=p_b, cov_type = "cov_equi", rho=0.5)

  # Generate linear predictor lp = X%*%beta where first 10 beta-coefficients are = a, all other = 0.
  lp <- generate_lp(X=X, p_nn=p_nn, a=1)

  # simulate gaussian response with mean = lp and sd = 1.
  Yg <- lp + rnorm(N)

  # simulate bernoulli response with mean = exp(lp)/(1+exp(lp)).
  Yb <- factor(rbinom(N, size=1, prob=exp(lp)/(1+exp(lp))))

  # simulate censored survival times from Cox regression with linear predictor lp:
  Tc <- simulWeib(N=N, lambda = 0.01, rho = 1, lp = lp)

  dat <- data.frame(Yg, Yb, Tc, X)

  return(dat)

}
```
This function simulates a toy data set with 30 covariates $X_1, \dots, X_{30}$, one continuous response $y_g$, one binary response $y_b$ and one set of censored survival times $T_c$. Let's now go through the individual components of the function one by one.

*Simulation of $X$:* The `generate_X` function simulates the rows of an $n \times p$ data frame $X$ independently from a multivariate Gaussian distribution with mean $0$ and $p \times p$ covariance matrix 
$$
\Sigma_{ij} = \left \{
\begin{array}{lr}
1\{i = j\}, & \text{Independent,} \\
\rho^{1\{i \neq j\}}, & \text{Equicorrelated,} \\
\rho^{|i-j|}, & \text{AR1},
\end{array}
\right.
$$
where $p_b$ randomly selected columns are then dichotomized with the indicator function $\delta(x)=1(x > 0)$.
```{r, eval=FALSE}
X <- generate_X(n=N, p=p, p_b=p_b, cov_type = "cov_equi", rho=0.5)
```
The covariance type is specified with the parameter `cov_type` and the correlation coefficient with `rho`. Each column of the resulting data.frame is either of class `"numeric"` (for the continuous columns) or `"factor"` (for the binary columns).

*Calculation of linear predictor:* The `generate_lp` function calculates the linear predictor $\ell_p=X\beta$ under sparsity,
where the first `p_nn` regression coefficients are non-zero, all other are set to zero. The (common) amplitude of the non-zero regression coefficients is specified with `a`. Here we generate $\ell_p$ that implies association with the first 10 covariates, each with amplitude $a=1$.
```{r, eval=FALSE}
lp <- generate_lp(X=X, p_nn=p_nn, a=1)
```
Note that inside `generate_lp` the model.matrix of `X` is first scaled. 

*Simulation of Gaussian response:* $Y_g$ is Gaussian with mean $\mu = \ell_p$ and standard deviation $\sigma = 1$:
```{r, eval=FALSE}
Yg <- lp + rnorm(N)
```

*Simulation of Bernoulli response:* $Y_b$ is Bernoulli with success probability $\mu = exp(\ell_p)/(1+exp(\ell_p))$:
```{r, eval=FALSE}
Yb <- factor(rbinom(N, size=1, prob=exp(lp)/(1+exp(lp))))
```

*Simulation of event and censoring times:* The final command:
```{r, eval=FALSE}
Tc <- simulWeib(N=N, lambda = 0.01, rho = 1, lp = lp)
```
generates censored survival times $T_c$ from a Cox regression model
$$
\lambda(t) = \lambda_0(t) \exp \left(\ell_p \right),
$$
with Weibull baseline hazard:
$$
\lambda_0(t) = \lambda_0 \rho t^{\rho-1}
$$
where $\lambda_0 > 0$ and $\rho > 0$ are scale and shape parameters, respectively. The censoring times $C$ are randomly drawn from an exponential distribution with a small (fixed) rate $\lambda_C=0.0005$, which results in very mild censoring. Once $T$ and $C$ have been simulated the function returns a survival object (`Surv`) with `time = min(T, C)` and `event = 1{T < C}`:
```{r}
simulWeib <- function(N, lambda0, rho, lp) {
    lambdaC = 5e-04
    v <- runif(n = N)
    Tlat <- (-log(v)/(lambda0 * exp(lp)))^(1/rho)
    C <- rexp(n = N, rate = lambdaC)
    time <- pmin(Tlat, C)
    status <- as.numeric(Tlat <= C)
    survival::Surv(time = time, event = status)
}
```

## The data set `simdata`

Now let's have a look at the first few columns of the data set `simdata`
```{r}
data(simdata)
```
```{r}
head(simdata[,1:9])
```
and finally confirm that `generate_simdata()` indeed reproduces the `simdata` dataset:
```{r}
all.equal(simdata, generate_simdata())
```
